---
title: 'Code for "Foggy connections: on the (non-)adaptation of lexical structures",
  Frontiers in Psychology'
author: "Matthias Urban"
date: "`r Sys.Date()`"
output: github_document
---

### Overview

This document provides code to run the analyses reported in the article "Foggy connections: on the (non-)adaptation of lexical structures". 

### Required datasets
• DEM model in file ""ETOPO_2022_v1_30s_N90W180_bed.tif""
  Available from  https://www.ncei.noaa.gov/products/etopo-global-relief-model (as of Nov 25)

• IDS data in files "forms.csv" and "languages.csv", version 4.2
  Available from the CLDF dataset for IDS at: https://zenodo.org/record/5547908

• South American dataset in file "cloudfog SA data and sources.tsv"
  Available from the Github repository for this article at: https://github.com/urban-m/cloudfog

### Required packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(ggplot2)
library(cowplot)
library(raster)
library(ggridges)
library(ggpubr)
```

### To begin, set seed for reproducibility

```{r}
set.seed(47)
```

### IDS analysis

#### Data preparation and exploration

###### Read in data 
```{r}
IDS_raw <- read.csv("forms.csv", header = T, encoding = "UTF-8")
IDS_languages <-read.csv("languages.csv", header = T, encoding = "UTF-8") 
```

##### Subset data to only keep relevant information, then join into a single dataset
```{r}
IDS_raw <- filter(IDS_raw, Parameter_ID == "1-730" | Parameter_ID == "1-740")
IDS_languages <- IDS_languages %>% rename(Language_ID = ID) %>% dplyr::select(Language_ID:Family)
IDS_raw <-left_join(IDS_raw, IDS_languages, by = "Language_ID", keep = FALSE) %>% 
    rename(lon=Longitude, lat=Latitude) %>% filter(!is.na(lon)) 
```

##### Check if languages are Colexifying or Distinguishing with regard to 'cloud' and 'fog'
```{r}
colex <- list()
for (i in unique(IDS_raw$Language_ID)){
  subset <- filter(IDS_raw, Language_ID == i)
  colex[i] <- ifelse(n_distinct(subset$Form) < nrow(subset), "Colexifying", "Distinguishing")
}
colex <- unlist(colex)
```

##### Now convert to a dataset in which there is one row per language, classified for the Colexification behavior
```{r}
IDS_data <- IDS_raw %>% group_by(Language_ID) %>% slice(1) %>% ungroup() %>%   
   dplyr::select(c(Glottolog_Name, Language_ID, Glottocode, Macroarea, lat, lon, Family)) %>% 
   mutate(Colex = colex)
```

##### Extract elevation from DEM, Log-transform elevation, and set Dutch to 1 masl
 ```{r, message=FALSE, warning=FALSE}
elevation_raster <- raster("ETOPO_2022_v1_30s_N90W180_bed.tif")
IDS_data <- mutate(IDS_data, Elevation = extract(elevation_raster, IDS_data[, c("lon", "lat")], 
   method="simple")) %>% mutate(IDS_data, LogElevation = log(Elevation))
IDS_data[74,"Elevation"] = 1
```

#### Create plots
 ```{r}
IDS_data_violinplot <- ggplot(IDS_data, aes(x = Colex, y = Elevation))  +
  geom_violin() +
  theme_minimal() +
  ggtitle("Whole IDS sample") +
  theme(plot.title = element_text(hjust = 0, size = 12, face = "bold"), legend.title =     
      element_blank(), legend.text = element_text(size = 12)) +
  theme(axis.title = element_text(size = 12)) +
  scale_x_discrete(labels = c("", "")) +
  geom_dotplot(aes(fill = Colex), color = NA,  binaxis = 'y', stackdir = 'center', binwidth = 65,       stackratio = 0.8, dotsize=1.2) +
  xlab("") +
  ylab("Elevation (masl)")
  
IDS_data_ridgeplot <- ggplot(IDS_data %>% group_by(Family) %>% filter(n() > 2), aes(x =                Elevation, y = fct_reorder(Family, Elevation, .desc = TRUE))) +
  theme_ridges() +
  ggtitle("Ridgeplots by family") +
  theme(plot.title = element_text(hjust = -1, , size = 12, face = "bold"), legend.title =     
      element_blank(), legend.text = element_text(size = 12)) +
  theme(axis.text=element_text(size=10)) +
   geom_density_ridges(
      aes(point_color = Colex, point_fill = Colex), jittered_points = TRUE, alpha = 0.3,        
      point_alpha = 1, point_size = 1.2, scale=0.85) +
  labs (x = "Elevation (masl)", y = "") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
  
IDS_data_panel <- ggarrange(IDS_data_violinplot, IDS_data_ridgeplot, nrow = 1,  common.legend = TRUE, legend = "bottom", font.label = list(size = 12))
IDS_data_panel
ggsave("IDS_data_panel.tiff", IDS_data_panel, dpi = 600)
```

#### Bayesian Mixed Effects Modelling

##### Set priors and build model
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
priors <- set_prior("normal(0, 2)", class = "b", coef = "LogElevation")
IDS_model <- brm(Colex ~ LogElevation + (1 + LogElevation | Family), family = "bernoulli", data =     IDS_data, seed = 10761, warmup = 8000, iter = 16000, chains = 4, control = list(adapt_delta =     0.999, max_treedepth = 20), save_pars = save_pars(all = TRUE), prior = priors)
```

##### check Rhat and ESS values
```{r}
summary(IDS_model)
```

##### Inspect chains
```{r}
plot(IDS_model)
```

##### Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(IDS_model)
pp_check(IDS_model, type = "error_binned")
```

##### Simpler model without Elevation as Predictor
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
IDS_nullmodel <- brm(Colex ~ 1 + (1 | Family), family = "bernoulli", data =   IDS_data, seed =  
    10761, warmup = 8000, iter = 16000, chains = 4, control = list(adapt_delta =    0.999,  
    max_treedepth = 20), save_pars = save_pars(all = TRUE))
```

##### Compute Bayes Factor
```{r, echo=FALSE, message=FALSE}
 bayes_factor(IDS_model, IDS_nullmodel)
```

### South American analysis

#### Data preparation and exploration

##### Read in data 
```{r}
SA_data <- read.csv("cloudfog SA data and sources.tsv", header = T, sep = "\t", encoding = "UTF-8")
```

##### Prepare data
```{r}
SA_data <- mutate(SA_data, Colex = as.character(Colex))
SA_data <- SA_data %>% rename(lon=Longitude, lat=Latitude)
```

##### Extract elevation from DEM, Log-transform elevation
 ```{r}
SA_data <- mutate(SA_data, Elevation = extract(elevation_raster, SA_data[, c("lon", "lat")], method="simple")) %>%  mutate(SA_data, LogElevation = log(Elevation))
```

##### Collapse families represented by only one language and isolates to one pseudo-group
 ```{r}
SA_data <- SA_data %>% group_by(Family) %>% mutate(Family2 = n()) 
SA_data <- SA_data %>% mutate(Family2 = replace(Family2, Family2 != 1, Family)) %>%            
  mutate(Family2 = replace(Family2, Family2 == 1, "Other families and isolates"))
```

##### Reduce data to a simple distinction of Colexifying and Distinguishing languages with respect to 'cloud' and 'fog'
 ```{r}
SA_databinary <- SA_data %>% filter(Colex != "Derivation") %>% filter(Colex!= "conflicting")
SA_databinary$Family2 <- factor(SA_databinary$Family2, levels = c("Arawakan", "Aymaran", 
   "Barbacoan", "Cahuapanan", "Huitotoan", "Jivaroan", "Matacoan", "Pano-Tacanan", "Quechuan",   
   "Tucanoan", "Tupian", "Zaparoan", "Other families and isolates"))
```

##### Log-transform elevation to reduce skew
```{r}
SA_data <- mutate(SA_data, LogElevation = log(Elevation))
```

##### Identify language families represented only by one language in the dataset (includes isolates) and assign these to one group
```{r}
SA_data <- SA_data %>% group_by(Family) %>% mutate(Family2 = n()) 
SA_data <- SA_data %>% mutate(Family2 = replace(Family2, Family2 != 1, Family)) %>% 
    mutate(Family2 = replace(Family2, Family2 == 1, "Other families and isolates"))
```

##### Reduce coding to a binary distinction
```{r}
SA_databinary <- SA_data %>% filter(Colex != "Derivation") %>% filter(Colex != "conflicting")
SA_databinary$Value <- ifelse(SA_databinary$Colex == "Colexifying", TRUE, FALSE)
```

#### Create plots
```{r}
SA_data_violinplot <- ggplot(SA_databinary, aes(x = Colex, y = Elevation))  +
  geom_violin() +
  theme_minimal() +
  ggtitle("Whole South American sample") +
  theme(plot.title = element_text(hjust = 0, size = 12, face = "bold"), legend.title =      
  element_blank(), legend.text = element_text(size = 12)) +
  theme(axis.title = element_text(size = 12)) +
  scale_x_discrete(labels = c("", "")) +
  geom_dotplot(aes(fill = Colex), color = NA,  binaxis = 'y', stackdir = 'center', binwidth = 65,       stackratio = 0.8, dotsize=1.2) +
  xlab("") +
  ylab("Elevation (masl)")
SA_data_violinplot 

SA_data_ridgeplot <- ggplot(SA_databinary %>% group_by(Family) %>% filter(n() > 2), 
  aes(x = Elevation, y = fct_reorder(Family, Elevation, .desc = TRUE))) +
  theme_ridges() +
  ggtitle("Ridgeplots by family") +
  theme(plot.title = element_text(hjust = -1), legend.title = element_blank(), legend.text =   
      element_text(size = 12)) +
  theme(axis.text=element_text(size=10)) +
  geom_density_ridges(
  aes(point_color = Colex, point_fill = Colex), jittered_points = TRUE, alpha = 0.3, point_alpha 
     = 1, point_size = 1.3, scale=0.85) +
  geom_point(data = subset(SA_databinary, Family == "Aymaran"), aes(x = Elevation), position =   
  position_nudge(y = 0.2), size = 1.2, color = "cyan") +
  labs (x = "Elevation (masl)", y = "") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
SA_data_ridgeplot
  
SA_data_panel <- ggarrange(SA_data_violinplot, SA_data_ridgeplot, nrow = 1,  common.legend = TRUE, legend = "bottom", font.label = list(size = 12))
SA_data_panel
ggsave("SA_data_panel.tiff", SA_data_panel, dpi = 600)
```

#### Bayesian Mixed Effects Modelling

###### Set priors and build model
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
priors <- set_prior("normal(0, 2)", class = "b", coef = "LogElevation")
SA_databinarymodel <- brm(Colex ~ LogElevation + (1 + LogElevation | Family2), family =     
   "bernoulli", data = SA_databinary, seed = 10761, warmup = 8000, iter = 12000, chains = 4,  
    control = list(adapt_delta = 0.999, max_treedepth = 20), prior = priors, save_pars =  
    save_pars(all = TRUE))
```

##### check Rhat and ESS values
```{r}
summary(SA_databinarymodel)
```

##### Inspect chains
```{r}
plot(SA_databinarymodel)
```

##### Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(SA_databinarymodel)
pp_check(SA_databinarymodel, type = "error_binned")
```

##### Simpler model without Elevation as Predictor
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
SA_databinarynnullmodel <- brm(Colex ~ 1 + (1 | Family), family = "bernoulli", data = 
   SA_databinary, seed = 10761, warmup = 8000, iter = 16000, chains = 4, control = 
   list(adapt_delta =  0.999, max_treedepth = 20), save_pars = save_pars(all = TRUE))
```

##### Compute Bayes Factor
```{r, echo=FALSE, message=FALSE}
 bayes_factor(SA_databinarymodel, SA_databinarynnullmodel)
```


#### Alternative model without isolates and singleton representatives of families collapsed
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
priors <- set_prior("normal(0, 2)", class = "b", coef = "LogElevation")
SA_databinarymodelalt <- brm(Colex ~ LogElevation + (1 + LogElevation | Family), family = 
   "bernoulli", data = SA_databinary, seed = 10761, warmup = 8000, iter = 16000, chains = 4, 
   control = list(adapt_delta = 0.999, max_treedepth = 20), prior = priors)
```

##### check Rhat and ESS values
```{r}
summary(SA_databinarymodelalt)
```

##### Inspect chains
```{r}
plot(SA_databinarymodelalt)
```

##### Inspect plots of observed data and posterior predictive samples
```{r, warning=FALSE}
pp_check(SA_databinarymodelalt)
pp_check(SA_databinarymodelalt, type = "error_binned")
```

#### Resampling and randomization test

##### Duplicate dataset and shuffle colexification values
```{r}
SA_databinary_shuffle<-SA_databinary
SA_databinary_shuffle$Value <- sample(SA_databinary_shuffle$Value, size = nrow(SA_databinary), replace = FALSE)
```

##### Procedure for resampling. This function adapted from Moran, Steven, Nicholas A. Lester and Eitan Grossman. 2021. Inferring recent evolutionary changes in speech sounds. Philosophical Transactions of the Royal Society B, 376. 20200198. https://doi.org/10.1098/rstb.2020.0198.
```{r}
resampling_withinFamilies =
  
  function(df){
    langs = vector()
    for(family_ID in unique(df$Family)){
      lang =
        sample(df$Name[df$Family == family_ID], 1)
      langs = c(langs, lang)
    }
    
    curr_sample = df[df$Name %in%
                       langs,]
    return(curr_sample)
  }
```


##### Resample. This is adapted from code by James S. Santangelo available at https://uoftcoders.github.io/rcourse/lec12-randomization-tests.html
```{r}
sim_means <- list()
resample_means <- list()

for(i in 1:10000){
  SA_databinary_resample<-resampling_withinFamilies(SA_databinary)
  SA_databinary_sim<-SA_databinary_resample
  SA_databinary_sim$Value <- sample(SA_databinary_sim$Value, size = nrow(SA_databinary_sim), 
      replace = FALSE)
  mean_identical_resample <- mean(SA_databinary_resample %>% filter(Value == TRUE) %>% 
     pull(Elevation))
  mean_distinct_resample <- mean(SA_databinary_resample %>% filter(Value == FALSE) %>%   
     pull(Elevation))
  mean_identical_sim <- mean(SA_databinary_sim %>% filter(Value == TRUE) %>% pull(Elevation))
  mean_distinct_sim <- mean(SA_databinary_sim %>% filter(Value == FALSE)  %>% pull(Elevation))
  mean_diff_resample <- mean_identical_resample - mean_distinct_resample
  mean_diff_sim <- mean_identical_sim - mean_distinct_sim
  resample_means[i] <- mean_diff_resample
  sim_means[i] <- mean_diff_sim
}
sim_means<-unlist(sim_means)
resample_means<-unlist(resample_means)
```

##### Plot results
```{r}
resampling_ggplot<- ggplot() +
  geom_histogram(aes(sim_means), fill ="cyan", binwidth = 20, color = "gray", alpha = 0.3) +
  geom_histogram(aes(resample_means), fill = "red", , binwidth = 20, color = "gray", alpha = 0.3) +
  xlab("Mean elevation of samples. Simulated data in cyan, observed data in red") +
  ylab("Number of observations") +
  theme_minimal()

resampling_ggplot
save_plot("resampling_boxplot.tiff", resampling_ggplot, base_width = 10, base_height = 6, dpi = 300)
```

##### Test for significance by the good old t-test.
```{r}
t.test(sim_means, resample_means)
```  

#### Generalized Linear Modelling for Arawakan

##### Filter data
```{r}
arawakan <- filter(SA_databinary, Family == 'Arawakan')
```

##### Build model
```{r}
arawakanmodel <- glm(Value ~ LogElevation, family = 'binomial', data = arawakan)
``` 

##### Check results
```{r}
summary(arawakanmodel)
``` 

##### Assess model visually
```{r}
plot(arawakanmodel)
``` 

##### Check overdispersion...not great but ok.
```{r}
pchisq(arawakanmodel$deviance, df=arawakanmodel$df.residual, lower.tail=FALSE)
``` 