---
title: 'Code for "Foggy connections: on the (non-)adaptation of lexical structures",
  Frontiers in Psychology'
author: "Matthias Urban"
date: "`r Sys.Date()`"
output: html_document
---

## Overview

This document provides code to run the analyses reported in the article "Foggy connections: on the (non-)adaptation of lexical structures". The dataset is provided in the GitHub repository SA_data under my account urban-m. This markdown file assumes that the file is located on the local machine in the working directory.

#### Required packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(ggplot2)
library(cowplot)
library(raster)
library(ggridges)
library(ggpubr)
```

#### Set seed for reproducibility

```{r}
set.seed(47)
```

## Data preparation and exploration

#### Read in data and convert to tibble
```{r}
elevation_raster <- raster("ETOPO_2022_v1_30s_N90W180_bed.tif")
SA_data <- read.csv("cloudfog SA data and sources.csv", header = T, sep = "\t", encoding = "UTF-8")
SA_data <- as_tibble(SA_data) %>% mutate(Value = as.character(Value))
```


SA_data <- SA_data %>% rename(lon=Longitude, lat=Latitude) 

SA_data <- mutate(SA_data, Elevation = extract(elevation_raster, SA_data[, c("lon", "lat")], method="simple"))


IDS_raw <- read.csv("forms.csv", header = T, encoding = "UTF-8")
IDS_raw <- filter(IDS_raw, Parameter_ID == "1-730" | Parameter_ID == "1-740")

IDS_languages <-read.csv("languages.csv", header = T, encoding = "UTF-8") 
IDS_languages <- IDS_languages %>% rename(Language_ID = ID)
IDS_languages <- IDS_languages %>% dplyr::select(Language_ID:Family)
IDS_raw <-left_join(IDS_raw, IDS_languages, by = "Language_ID", keep = FALSE) %>% rename(lon=Longitude, lat=Latitude) %>% filter(!is.na(lon)) 



colex <- list()
for (i in unique(IDS_raw$Language_ID)){
  subset <- filter(IDS_raw, Language_ID == i)
  colex[i] <- ifelse(n_distinct(subset$Form) < nrow(subset), "Colexifying", "Distinguishing")
}
colex <- unlist(colex)


 IDS_data <- IDS_raw %>% group_by(Language_ID) %>% slice(1) %>% ungroup() %>% dplyr::select(c(Glottolog_Name, Language_ID, Glottocode, Macroarea, lat, lon, Family)) %>% mutate(Colex = colex)
 
IDS_data <- mutate(IDS_data, Elevation = extract(elevation_raster, IDS_data[, c("lon", "lat")], method="simple"))
IDS_data[74,"Elevation"] = 1

tibble(colex)

https://www.ncei.noaa.gov/products/etopo-global-relief-model (as of Nov 25)



IDS_data_violinplot <- ggplot(IDS_data, aes(x = Colex, y = Elevation))  +
  geom_violin() +
  theme_minimal() +
  ggtitle("Whole IDS sample") +
  theme(plot.title = element_text(hjust = 0, size = 14, face = "bold"), legend.title = element_blank(), legend.text = element_text(size = 14)) +
  theme(axis.title = element_text(size = 14)) +
  scale_x_discrete(labels = c("", "")) +
  geom_dotplot(aes(fill = Colex), color = NA,  binaxis = 'y', stackdir = 'center', binwidth = 65, stackratio = 0.8, dotsize=0.8) +
  xlab("") +
  ylab("Elevation (masl)")


IDS_data <- mutate(IDS_data, LogElevation = log(Elevation))
priors <- set_prior("normal(0, 2)", class = "b", coef = "LogElevation")
SA_dataIDSmodel <- brm(Colex ~ LogElevation + (1 + LogElevation | Family), family = "bernoulli", data = IDS_data, seed = 10761, warmup = 8000, iter = 12000, chains = 4, control = list(adapt_delta = 0.999, max_treedepth = 20), prior = priors)

IDS_data_ridgeplot <- ggplot(IDS_data %>% group_by(Family) %>% filter(n() > 2), aes(x = Elevation, y = fct_reorder(Family, Elevation, .desc = TRUE))) +
  theme_ridges() +
  ggtitle("Ridgeplots by family") +
  theme(plot.title = element_text(hjust = -0.5), legend.title = element_blank(), legend.text = element_text(size = 14)) +
  theme(axis.text=element_text(size=10)) +
   geom_density_ridges(
    aes(point_color = Colex, point_fill = Colex), jittered_points = TRUE, alpha = 0.3, point_alpha = 1, point_size = 2, scale=0.85) +
  labs (x = "Elevation (masl)", y = "") +
  scale_x_continuous(expand = c(0, 0)) +
   scale_y_discrete(expand = c(0, 0))

IDS_data_panel <- ggarrange(IDS_data_violinplot, IDS_data_ridgeplot, nrow = 1,  common.legend = TRUE, legend = "bottom", font.label = list(size = 14))
IDS_data_panel
ggsave("densityplotpanel.pdf", densityplotpanel, dpi = 600)

#### Log-transform elevation to reduce skew
```{r}
SA_data <- mutate(SA_data, LogElevation = log(Elevation))
```

#### Identify language families represented only by one language in the dataset (includes isolates) and assign these to one group
```{r}
SA_data <- SA_data %>% group_by(Family) %>% mutate(Family2 = n()) 
SA_data <- SA_data %>% mutate(Family2 = replace(Family2, Family2 != 1, Family)) %>% mutate(Family2 = replace(Family2, Family2 == 1, "Other families and isolates"))
```

#### Reduce coding to a binary distinction
```{r}
SA_databinary <- SA_data %>% filter(Value != "Derivation") %>% filter(Value != "conflicting")
SA_databinary$Value <- ifelse(SA_databinary$Value == "Colexifying", TRUE, FALSE)
```

#### Create plot of distributions
```{r}
SA_databinary$Family2 <- factor(SA_databinary$Family2, levels = c("Arawakan", "Aymaran", "Barbacoan", "Cahuapanan", "Huitotoan", "Jivaroan", "Matacoan", "Pano-Tacanan", "Quechuan", "Tucanoan", "Tupian", "Zaparoan", "Other families and isolates"))
SA_data_boxplot <- ggplot(SA_databinary, aes(x = Value, y = Elevation))  +
  geom_violin() +
  scale_fill_manual(values = c("#000000", "#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#998040", "#0072B2", "#a7a5ec", "#D55E00", "#CC79A7", "#9999CC", "#FFFFFF")) +
  geom_dotplot(aes(fill = Family2), binaxis = 'y', stackdir = 'center', binwidth = 65, stackratio = 0.8, dotsize=1.2)  +
  labs(fill = "Family") +
  scale_x_discrete(labels = c("Distinct terms", "Colexification of 'cloud' and 'fog'")) +
  xlab("") +
  ylab("Elevation (masl)")
SA_data_boxplot
save_plot("SA_data_boxplot.tiff", SA_data_boxplot, base_width = 10, base_height = 6, dpi = 300)
```

# Bayesian Mixed Effects Modelling

#### Set priors and build model
```{r, echo=FALSE, results='hide'}
priors <- set_prior("normal(0, 2)", class = "b", coef = "LogElevation")
SA_databinarymodel <- brm(Value ~ LogElevation + (1 + LogElevation | Family2), family = "bernoulli", data = SA_databinary, seed = 10761, warmup = 8000, iter = 12000, chains = 4, control = list(adapt_delta = 0.999, max_treedepth = 20), prior = priors)
```

##### check Rhat and ESS values
```{r}
summary(SA_databinarymodel)
```

##### Inspect chains
```{r}
plot(SA_databinarymodel)
```

##### Inspect plots of observed data and posterior predictive samples
```{r}
pp_check(SA_databinarymodel)
pp_check(SA_databinarymodel, type = "error_binned")
```

##### Assess posterior probability versus chance
```{r}
SA_databinarymodelsamples <- as_draws_df(SA_databinarymodel)
sum(SA_databinarymodelsamples$b_LogElevation < 0) / nrow(SA_databinarymodelsamples)
```

#### Alternative model without isolates and singleton representatives of families collapsed
```{r, echo=FALSE, results='hide'}
priors <- set_prior("normal(0, 2)", class = "b", coef = "LogElevation")
SA_databinarymodelalt <- brm(Value ~ LogElevation + (1 + LogElevation | Family), family = "bernoulli", data = SA_databinary, seed = 10761, warmup = 8000, iter = 12000, chains = 4, control = list(adapt_delta = 0.999, max_treedepth = 20), prior = priors)
```

##### Inspect chains
```{r}
plot(SA_databinarymodelalt)
```

##### Inspect plots of observed data and posterior predictive samples
```{r, warning=FALSE}
pp_check(SA_databinarymodelalt)
pp_check(SA_databinarymodelalt, type = "error_binned")
```

##### Assess posterior probability versus chance
```{r, warning=FALSE}
SA_databinarymodelaltsamples <- as_draws_df(SA_databinarymodelalt)
sum(SA_databinarymodelaltsamples$b_LogElevation < 0) / nrow(SA_databinarymodelaltsamples)
```

# Resampling and randomization test

#### Duplicate dataset and shuffle colexification values
```{r}
SA_databinary_shuffle<-SA_databinary
SA_databinary_shuffle$Value <- sample(SA_databinary_shuffle$Value, size = nrow(SA_databinary), replace = FALSE)
```

#### Procedure for resampling. This function adapted from 
Moran, Steven, Nicholas A. Lester and Eitan Grossman. 2021. Inferring recent evolutionary changes in speech sounds. Philosophical Transactions of the Royal Society B, 376. 20200198. https://doi.org/10.1098/rstb.2020.0198.
```{r}
resampling_withinFamilies =
  
  function(df){
    langs = vector()
    for(family_ID in unique(df$Family)){
      lang =
        sample(df$Name[df$Family == family_ID], 1)
      langs = c(langs, lang)
    }
    
    curr_sample = df[df$Name %in%
                       langs,]
    return(curr_sample)
  }
```


#### Resample. This is adapted from code by James S. Santangelo available at https://uoftcoders.github.io/rcourse/lec12-randomization-tests.html
```{r}
sim_means <- list()
resample_means <- list()

for(i in 1:10000){
  SA_databinary_resample<-resampling_withinFamilies(SA_databinary)
  SA_databinary_sim<-SA_databinary_resample
  SA_databinary_sim$Value <- sample(SA_databinary_sim$Value, size = nrow(SA_databinary_sim), replace = FALSE)
  mean_identical_resample <- mean(SA_databinary_resample %>% filter(Value == TRUE) %>% pull(Elevation))
  mean_distinct_resample <- mean(SA_databinary_resample %>% filter(Value == FALSE) %>% pull(Elevation))
  mean_identical_sim <- mean(SA_databinary_sim %>% filter(Value == TRUE) %>% pull(Elevation))
  mean_distinct_sim <- mean(SA_databinary_sim %>% filter(Value == FALSE)  %>% pull(Elevation))
  mean_diff_resample <- mean_identical_resample - mean_distinct_resample
  mean_diff_sim <- mean_identical_sim - mean_distinct_sim
  resample_means[i] <- mean_diff_resample
  sim_means[i] <- mean_diff_sim
}
sim_means<-unlist(sim_means)
resample_means<-unlist(resample_means)
```

#### Plot results
```{r}
resampling_ggplot<- ggplot() +
  geom_histogram(aes(sim_means), fill ="cyan", binwidth = 20, color = "gray", alpha = 0.3) +
  geom_histogram(aes(resample_means), fill = "red", , binwidth = 20, color = "gray", alpha = 0.3) +
  xlab("Mean elevation of samples. Simulated data in cyan, observed data in red") +
  ylab("Number of observations")

resampling_ggplot
save_plot("resampling_boxplot.tiff", resampling_ggplot, base_width = 10, base_height = 6, dpi = 300)
```

#### Test for significance by the good old t-test.
```{r}
t.test(sim_means, resample_means)
```  

# Generalized Linear Modelling for Arawakan

#### Filter data
```{r}
arawakan <- filter(SA_databinary, Family == 'Arawakan')
```

#### Build model
```{r}
arawakanmodel <- glm(Value ~ LogElevation, family = 'binomial', data = arawakan)
``` 

#### Check results
```{r}
summary(arawakanmodel)
``` 

#### Assess model visually
```{r}
plot(arawakanmodel)
``` 

#### Check overdispersion...not great but ok.
```{r}
pchisq(arawakanmodel$deviance, df=arawakanmodel$df.residual, lower.tail=FALSE)
``` 